---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 4,
  dpi = 150
)
```

# ggaib

ggplot2 themes and color scales for the Annenberg Institute at Brown University.

## Installation

```r
# install.packages("devtools")
devtools::install_github("andrewmcamp/ggaib")
```

## Brand Colors

```{r colors}
library(ggaib)
library(ggplot2)

aib_colors()
```

## Themes

### `theme_aib()` — Publication

A minimal, publication-quality theme with no gridlines and clean axis lines.

```{r theme-aib}
ggplot(mtcars, aes(mpg, wt)) +
  geom_point(aes(color = factor(cyl)), size = 3) +
  scale_color_aib() +
  labs(
    title = "Vehicle Weight by Fuel Efficiency",
    subtitle = "Motor Trend Car Road Tests, 1974",
    caption = "Source: mtcars dataset"
  ) +
  theme_aib()
```

### `theme_aib_grid()` — Data-Dense

Same as `theme_aib()` with light gray major gridlines for data-dense plots.

```{r theme-grid}
ggplot(economics, aes(date, unemploy / 1000)) +
  geom_line(color = aib_colors("navy"), linewidth = 0.7) +
  labs(
    title = "US Unemployment Over Time",
    subtitle = "In thousands",
    y = "Unemployed (thousands)",
    x = NULL
  ) +
  theme_aib_grid()
```

### `theme_aib_slide()` — Presentations

Larger text and more spacing, optimized for slides.

```{r theme-slide}
ggplot(mtcars, aes(factor(cyl), fill = factor(cyl))) +
  geom_bar() +
  scale_fill_aib() +
  labs(title = "Cars by Cylinder Count") +
  theme_aib_slide()
```

### `theme_aib_map()` — Maps

No axes, ticks, or gridlines. Retains title and caption styling.

```{r theme-map}
ggplot(mtcars, aes(mpg, wt)) +
  geom_point(color = aib_colors("navy"), size = 3) +
  labs(title = "Vehicle Weight by Fuel Efficiency") +
  theme_aib_map()
```

## Color Scales

### Continuous

```{r scale-continuous}
ggplot(mtcars, aes(mpg, wt, color = hp)) +
  geom_point(size = 3) +
  scale_color_aib_c() +
  labs(title = "Horsepower Gradient", color = "HP") +
  theme_aib()
```

### Diverging

```{r scale-diverging}
mtcars$hp_centered <- mtcars$hp - mean(mtcars$hp)

ggplot(mtcars, aes(mpg, wt, color = hp_centered)) +
  geom_point(size = 3) +
  scale_color_aib_div() +
  labs(title = "Horsepower (Centered)", color = "HP - Mean") +
  theme_aib()
```
