---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "ragg_png",
  fig.width = 7,
  fig.height = 5,
  dpi = 600,
  out.width = "672"
)
```

# ggaib

ggplot2 themes and color scales for the Annenberg Institute at Brown University.

## Installation

```r
# install.packages("devtools")
devtools::install_github("andrewmcamp/ggaib")
```

## Brand Colors

```{r colors}
library(ggaib)
library(ggplot2)

aib_colors()
```

## Themes
There are four core themes included in the `ggaib` pacakge, each optimized for a different use case. All themes share the same base font and color palette, but differ in gridline and axis styling to suit different types of visualizations.

### `theme_aib()` — Publication

A minimal, publication-quality theme with no gridlines and clean axis lines.

```{r theme-aib, fig.alt = "Scatter plot of per-pupil spending versus math scores by urbanicity using theme_aib"}
set.seed(42)
districts <- data.frame(
  spending = c(rnorm(40, 11, 2), rnorm(40, 15, 2.5), rnorm(40, 13, 3)),
  avg_score = c(rnorm(40, 250, 15), rnorm(40, 270, 12), rnorm(40, 258, 18)),
  urbanicity = rep(c("Urban", "Suburban", "Rural"), each = 40)
)

ggplot(districts, aes(spending, avg_score, color = urbanicity)) +
  geom_point(size = 2) +
  scale_color_aib() +
  scale_x_continuous(labels = aib_label("dollar")) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Per-Pupil Spending and Math Achievement",
    subtitle = "Simulated school district data",
    x = "Per-Pupil Expenditures\n($1,000s)",
    y = "Avgerage Math Score",
    color = "Urbanicity",
    caption = "Note: Simulated data for illustration"
  ) +
  theme_aib()
```

### `theme_aib_grid()` — Data-Dense

Same as `theme_aib()` with light gray major gridlines for data-dense plots.

```{r theme-grid, fig.alt = "Line chart showing a narrowing achievement gap between higher-income and lower-income students over time using theme_aib_grid"}
set.seed(42)
years <- 2005:2023
gap_data <- data.frame(
  year = rep(years, 2),
  group = rep(c("Higher-Income", "Lower-Income"), each = length(years)),
  score = c(
    270 + cumsum(rnorm(length(years), 0.3, 0.8)),
    240 + cumsum(rnorm(length(years), 0.8, 0.9))
  )
)

ggplot(gap_data, aes(year, score, color = group)) +
  geom_line(linewidth = 1) +
  scale_color_aib() +
  scale_x_continuous(
    breaks = seq(2005, 2025, 3)
  ) +
  scale_y_continuous(
    limits = c(215, 300),
    breaks = seq(200, 300, 20)
  ) +
  labs(
    title = "Reading Achievement Gap Over Time",
    subtitle = "4th-grade scores by household income",
    x = NULL,
    y = "Average Reading Score",
    color = NULL
  ) +
  theme_aib_grid()
```

### `theme_aib_slide()` — Presentations

Larger text and more spacing, optimized for slides.

```{r theme-slide, fig.width = 8.75, fig.height = 6.25, fig.alt = "Bar chart of K-12 enrollment by school type using theme_aib_slide"}
enrollment <- data.frame(
  type = c("Traditional\nPublic", "Charter", "Magnet", "Private"),
  students = c(47.3, 3.7, 2.5, 5.7)
)
enrollment$type <- factor(enrollment$type, levels = enrollment$type)
enrollment$students <- enrollment$students * 1e6

ggplot(enrollment, aes(type, students, fill = type)) +
  geom_col() +
  scale_fill_aib() +
  scale_y_continuous(labels = aib_label("comma")) +
  labs(
    title = "U.S. K\u201312 Enrollment by School Type",
    x = "Sector",
    y = "Enrollments"
  ) +
  theme_aib_slide() +
  theme(legend.position = "none")
```

### `theme_aib_map()` — Maps

No axes, ticks, or gridlines. Retains title and caption styling.

```{r theme-map, fig.alt = "Choropleth map of per-pupil spending by U.S. state using theme_aib_map"}
set.seed(42)
states <- ggplot2::map_data("state")
spending_by_state <- data.frame(
  region = unique(states$region),
  spending = runif(length(unique(states$region)), 7, 24)
)
map_df <- merge(states, spending_by_state, by = "region")

ggplot(map_df, aes(long, lat, group = group, fill = spending)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  scale_fill_aib_c() +
  labs(
    title = "Per-Pupil Spending by State",
    fill = "$ (1,000s)",
    caption = "Note: Simulated data for illustration"
  ) +
  coord_fixed(1.3) +
  theme_aib_map()
```

## Color Scales

### Continuous

```{r scale-continuous, fig.alt = "Scatter plot of schools colored by student-teacher ratio using a continuous navy-to-sky gradient"}
set.seed(42)
schools <- data.frame(
  enrollment = runif(80, 200, 1500),
  avg_score = rnorm(80, 260, 20),
  st_ratio = runif(80, 10, 30)
)
schools$avg_score <- schools$avg_score - (schools$st_ratio - 20) * 1.5

ggplot(schools, aes(enrollment, avg_score, color = st_ratio)) +
  geom_point(size = 2) +
  scale_color_aib_c() +
  scale_x_continuous(labels = aib_label("comma"), breaks = seq(0, 1500, 300)) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Student-Teacher Ratio Across Schools",
    x = "Enrollment",
    y = "Avg. Test Score",
    color = "Student-Teacher\nRatio"
  ) +
  theme_aib()
```

### Diverging

```{r scale-diverging, fig.alt = "Scatter plot of districts colored by year-over-year spending change using a diverging red-gray-navy scale"}
set.seed(42)
districts2 <- data.frame(
  enrollment = runif(80, 500, 5000),
  avg_score = rnorm(80, 255, 20),
  spending_change = rnorm(80, 0, 6)
)

ggplot(districts2, aes(enrollment, avg_score, color = spending_change)) +
  geom_point(size = 2) +
  scale_color_aib_div() +
  scale_x_continuous(labels = aib_label("comma"), breaks = seq(0, 5000, 500)) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Year-over-Year Spending Change",
    x = "Enrollment",
    y = "Avg. Test Score",
    color = "Spending\nChange (%)"
  ) +
  theme_aib()
```
