---
title: "Font System"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Font System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The ggaib package manages fonts automatically so that plots render
consistently across machines. When you load the package, it selects the
best available font for each of three typographic roles --- heading,
body, and serif --- and tells you what it found.

## Font roles

Every `theme_aib()` variant assigns fonts to plot elements through three
roles:

| Role | Used for | Brand font | Bundled alternative |
|:----------|:-------------------------------|:-------------------|:--------------------|
| `heading` | Plot title, subtitle, facet strips | Area Normal | Albert Sans |
| `body` | Axis labels, axis titles, legend | Source Sans Pro | Source Sans 3 |
| `serif` | Plot caption | Minion Pro | Crimson Text |

Retrieve the active family name for any role with `aib_font()`:

```{r}
library(ggaib)

aib_font("heading")
aib_font("body")
aib_font("serif")
```

Or retrieve all three at once:

```{r}
str(aib_font_families())
```

## How fonts are selected

When the package loads it works through a three-tier priority list for
each role:

1. **Brand font** --- If the preferred font (e.g. Area Normal) is
   installed on the system, it is used directly.
2. **Bundled alternative** --- If the brand font is not found, ggaib
   registers an open-source substitute shipped in the package's
   `inst/fonts` directory using `systemfonts::register_font()`. Each
   alternative is bundled in four weights (regular, italic, bold, and
   bold italic).
3. **System default** --- If font registration fails entirely (for
   example, if systemfonts is unavailable), the package falls back to
   the generic `"sans"` and `"serif"` families.

Because the bundled fonts are included in the package itself, the
fallback works offline and does not require downloading anything at
runtime.

## Startup messages

When ggaib is attached it prints a single line indicating which tier is
active. The message corresponds to one of four scenarios:

```
## All brand fonts installed
Using AIB brand fonts (Area Normal, Source Sans Pro, Minion Pro).

## No brand fonts; bundled alternatives registered
Brand fonts not found. Using bundled alternatives (Albert Sans, Source Sans 3, Crimson Text).

## Font registration failed
Font registration failed. Using system defaults (sans/serif).

## Mixed (some brand, some bundled)
Fonts: Area Normal (brand), Source Sans 3 (bundled), Crimson Text (bundled).
```

If you see the "bundled alternatives" message, plots will still look
consistent --- the open-source fonts were chosen to be close
typographic matches for the brand typefaces.

## Graphics device

The font system relies on the
[systemfonts](https://github.com/r-lib/systemfonts) package to register
bundled fonts. For these registered fonts to render correctly, use a
graphics device that supports systemfonts, such as the
[ragg](https://ragg.r-lib.org) package:

```r
# In an R script
ragg::agg_png("plot.png", width = 7, height = 5, units = "in", res = 300)
print(p)
dev.off()
```

RStudio and Quarto use ragg by default when it is installed. In an
RMarkdown document, set the device in the chunk options:

````
```{r, dev = "ragg_png"}`r ''`
```
````
