---
title: "Getting Started with ggaib"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ggaib}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  fig.width = 3.5,
  fig.height = 2,
  dpi = 600
)
showtext::showtext_opts(dpi = 600)
```

## Overview

The `ggaib` package provides ggplot2 themes and color scales for the
Annenberg Institute at Brown University. It includes four theme variants
and discrete, continuous, and diverging color scales using the Institute's
brand palette.

```{r library}
library(ggaib)
library(ggplot2)
```

## Brand Colors

Access the full palette or select specific colors:

```{r colors}
aib_colors()
aib_colors("navy", "red", "emerald")
```

## Themes
There are four core themes included in the `ggaib` pacakge, each optimized for a different use case. All themes share the same base font and color palette, but differ in gridline and axis styling to suit different types of visualizations.

### Publication (default)

```{r theme-base}
set.seed(42)
districts <- data.frame(
  spending = c(rnorm(40, 11, 2), rnorm(40, 15, 2.5), rnorm(40, 13, 3)),
  avg_score = c(rnorm(40, 250, 15), rnorm(40, 270, 12), rnorm(40, 258, 18)),
  urbanicity = rep(c("Urban", "Suburban", "Rural"), each = 40)
)

ggplot(districts, aes(spending, avg_score, color = urbanicity)) +
  geom_point(size = 2) +
  scale_color_aib() +
  scale_x_continuous(labels = aib_label("dollar")) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Per-Pupil Spending and Math Achievement",
    subtitle = "Simulated school district data",
    x = "Per-Pupil Expenditures\n($1,000s)",
    y = "Avgerage Math Score",
    color = "Urbanicity",
    caption = "Note: Simulated data for illustration"
  ) +
  theme_aib()
```

### Grid

```{r theme-grid}
set.seed(42)
years <- 2005:2023
gap_data <- data.frame(
  year = rep(years, 2),
  group = rep(c("Higher-Income", "Lower-Income"), each = length(years)),
  score = c(
    270 + cumsum(rnorm(length(years), 0.3, 0.8)),
    240 + cumsum(rnorm(length(years), 0.8, 0.9))
  )
)

ggplot(gap_data, aes(year, score, color = group)) +
  geom_line(linewidth = 1) +
  scale_color_aib() +
  scale_x_continuous(
    breaks = seq(2005, 2025, 3)
  ) +
  scale_y_continuous(
    limits = c(215, 300),
    breaks = seq(200, 300, 20)
  ) +
  labs(
    title = "Reading Achievement Gap Over Time",
    subtitle = "4th-grade scores by household income",
    x = NULL,
    y = "Average Reading Score",
    color = NULL
  ) +
  theme_aib_grid()
```

### Slide

```{r theme-slide}
enrollment <- data.frame(
  type = c("Traditional\nPublic", "Charter", "Magnet", "Private"),
  students = c(47.3, 3.7, 2.5, 5.7)
)
enrollment$type <- factor(enrollment$type, levels = enrollment$type)
enrollment$students <- enrollment$students * 1e6

ggplot(enrollment, aes(type, students, fill = type)) +
  geom_col() +
  scale_fill_aib() +
  scale_y_continuous(labels = aib_label("comma")) +
  labs(
    title = "U.S. K\u201312 Enrollment by School Type",
    x = "Sector",
    y = "Enrollments"
  ) +
  theme_aib_slide() +
  theme(legend.position = "none")
```

### Map

```{r theme-map}
set.seed(42)
states <- ggplot2::map_data("state")
spending_by_state <- data.frame(
  region = unique(states$region),
  spending = runif(length(unique(states$region)), 7, 24)
)
map_df <- merge(states, spending_by_state, by = "region")

ggplot(map_df, aes(long, lat, group = group, fill = spending)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  scale_fill_aib_c() +
  labs(
    title = "Per-Pupil Spending by State",
    fill = "$ (1,000s)",
    caption = "Note: Simulated data for illustration"
  ) +
  coord_fixed(1.3) +
  theme_aib_map()
```

## Color Scales

### Continuous

```{r scale-continuous}
set.seed(42)
schools <- data.frame(
  enrollment = runif(80, 200, 1500),
  avg_score = rnorm(80, 260, 20),
  st_ratio = runif(80, 10, 30)
)
schools$avg_score <- schools$avg_score - (schools$st_ratio - 20) * 1.5

ggplot(schools, aes(enrollment, avg_score, color = st_ratio)) +
  geom_point(size = 2) +
  scale_color_aib_c() +
  scale_x_continuous(labels = aib_label("comma"), breaks = seq(0, 1500, 300)) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Student-Teacher Ratio Across Schools",
    x = "Enrollment",
    y = "Avg. Test Score",
    color = "Student-Teacher\nRatio"
  ) +
  theme_aib()
```

### Diverging

```{r scale-diverging}
set.seed(42)
districts2 <- data.frame(
  enrollment = runif(80, 500, 5000),
  avg_score = rnorm(80, 255, 20),
  spending_change = rnorm(80, 0, 6)
)

ggplot(districts2, aes(enrollment, avg_score, color = spending_change)) +
  geom_point(size = 2) +
  scale_color_aib_div() +
  scale_x_continuous(labels = aib_label("comma"), breaks = seq(0, 5000, 500)) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Year-over-Year Spending Change",
    x = "Enrollment",
    y = "Avg. Test Score",
    color = "Spending\nChange (%)"
  ) +
  theme_aib()
```
