---
title: "Axis Labels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Axis Labels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  fig.width = 7,
  fig.height = 5,
  dpi = 600
)
```

```{r library}
library(ggaib)
library(ggplot2)
```

`aib_label()` is a convenience wrapper around common
[scales](https://scales.r-lib.org) label functions. It returns a
labeler function that can be passed directly to the `labels` argument of
any continuous scale.

## Available formats

| Type | Output | Underlying function |
|:----------|:-----------|:--------------------------------|
| `"comma"` | 1,000 | `scales::label_comma()` |
| `"dollar"` | $1,000 | `scales::label_dollar()` |
| `"percent"` | 100% | `scales::label_percent()` |
| `"pp"` | 1 pp | `scales::label_comma(suffix = " pp")` |

## Examples

### Comma

```{r comma}
enrollment <- data.frame(
  sector = c("Public", "Charter", "Magnet", "Private"),
  students = c(47.3, 3.7, 2.5, 5.7) * 1e6
)
enrollment$sector <- factor(enrollment$sector, levels = enrollment$sector)

ggplot(enrollment, aes(sector, students, fill = sector)) +
  geom_col() +
  scale_fill_aib() +
  scale_y_continuous(labels = aib_label("comma")) +
  labs(title = "Comma Format", x = NULL, y = "Students") +
  theme_aib() +
  theme(legend.position = "none")
```

### Dollar

```{r dollar}
spending <- data.frame(
  state = c("NY", "NJ", "CT", "MA", "VT"),
  ppexp = c(26.6, 21.0, 21.8, 18.6, 20.3) * 1000
)
spending$state <- factor(spending$state, levels = spending$state)

ggplot(spending, aes(state, ppexp, fill = state)) +
  geom_col() +
  scale_fill_aib() +
  scale_y_continuous(labels = aib_label("dollar")) +
  labs(title = "Dollar Format", x = NULL, y = "Per-Pupil Spending") +
  theme_aib() +
  theme(legend.position = "none")
```

### Percent

```{r percent}
grad <- data.frame(
  year = 2015:2023,
  rate = seq(0.83, 0.87, length.out = 9) + rnorm(9, 0, 0.005)
)

ggplot(grad, aes(year, rate)) +
  geom_line(linewidth = 1, color = aib_colors("navy")) +
  geom_point(size = 2, color = aib_colors("navy")) +
  scale_y_continuous(labels = aib_label("percent")) +
  labs(title = "Percent Format", x = NULL, y = "Graduation Rate") +
  theme_aib_grid()
```

### Percentage points

```{r pp}
set.seed(42)
effects <- data.frame(
  program = c("Tutoring", "Summer School", "Mentoring", "After-School"),
  effect = c(3.2, 1.8, 2.5, 1.1)
)
effects$program <- factor(effects$program, levels = effects$program)

ggplot(effects, aes(program, effect, fill = program)) +
  geom_col() +
  scale_fill_aib() +
  scale_y_continuous(labels = aib_label("pp")) +
  labs(
    title = "Percentage Point Format",
    x = NULL,
    y = "Effect on Graduation Rate"
  ) +
  theme_aib() +
  theme(legend.position = "none")
```

## Usage

Pass `aib_label()` to the `labels` argument of any continuous scale:
```r
scale_x_continuous(labels = aib_label("dollar"))
scale_y_continuous(labels = aib_label("percent"))
```

For formats not covered by `aib_label()`, use the
[scales](https://scales.r-lib.org) package directly.
