---
title: "Brand Palette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brand Palette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  fig.width = 7,
  fig.height = 5,
  dpi = 600
)
```

```{r library}
library(ggaib)
library(ggplot2)
```

The AIB brand palette contains eight colors. Two functions provide
access to these colors for use in custom plots or analyses.

## `aib_colors()` --- Access colors by name

Call `aib_colors()` with no arguments to see the full palette:

```{r all-colors}
aib_colors()
```

Pass color names to select specific values:

```{r select-colors}
aib_colors("navy", "red", "emerald")
```

An error is raised for invalid names:

```{r invalid-color, error = TRUE}
aib_colors("purple")
```

### Palette swatch

```{r swatch, fig.height = 1.5}
cols <- aib_colors()
df <- data.frame(
  name = factor(names(cols), levels = names(cols)),
  y = 1
)

ggplot(df, aes(name, y, fill = name)) +
  geom_tile(color = "white", linewidth = 1) +
  scale_fill_manual(values = cols) +
  labs(title = "AIB Brand Palette") +
  theme_aib() +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none"
  )
```

### Using colors directly

`aib_colors()` returns standard hex codes, so the values can be used
anywhere ggplot2 accepts a color string:

```{r direct-use}
set.seed(42)
df <- data.frame(x = 1:50, y = cumsum(rnorm(50)))

ggplot(df, aes(x, y)) +
  geom_line(linewidth = 1, color = aib_colors("navy")) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = aib_colors("gray")
  ) +
  labs(title = "Using Colors Directly", x = NULL, y = NULL) +
  theme_aib_grid()
```

## `aib_pal()` --- Palette function for scales

`aib_pal()` returns a function that generates `n` colors from the
palette. This is the engine behind `scale_color_aib()` and
`scale_fill_aib()`, but it can also be used on its own:

```{r pal-function}
pal <- aib_pal()
pal(3)
pal(5)
```

This is useful when you need brand colors outside of a ggplot2 scale,
for example when building a custom legend or annotating a plot with
colored text.

## Color reference

| Name | Hex | Preview |
|:---------|:----------|:---|
| navy | `#1B3E6F` | |
| red | `#C00404` | |
| emerald | `#00AF9A` | |
| yellow | `#EBA900` | |
| sky | `#55C8E8` | |
| taupe | `#B6B09D` | |
| brown | `#503629` | |
| gray | `#97A3AE` | |
