---
title: "Preparing Figures for Presentations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing Figures for Presentations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  fig.width = 7,
  fig.height = 5,
  dpi = 600
)
```

```{r library}
library(ggaib)
library(ggplot2)
```

Any ggaib theme can be adapted for slides by increasing the `base_size`
parameter. The default `base_size = 11` is tuned for print and on-screen
reading; bumping it to 16 or higher makes text legible when projected.

## Default vs. presentation sizing

```{r base-plot, include = FALSE}
set.seed(42)
enrollment <- data.frame(
  type = c("Traditional\nPublic", "Charter", "Magnet", "Private"),
  students = c(47.3, 3.7, 2.5, 5.7) * 1e6
)
enrollment$type <- factor(enrollment$type, levels = enrollment$type)
```

### `base_size = 11` (default)

```{r size-default}
ggplot(enrollment, aes(type, students, fill = type)) +
  geom_col() +
  scale_fill_aib() +
  scale_y_continuous(labels = aib_label("comma")) +
  labs(
    title = "U.S. K\u201312 Enrollment by Sector",
    x = NULL,
    y = "Students"
  ) +
  theme_aib() +
  theme(legend.position = "none")
```

### `base_size = 16` (presentation)

```{r size-presentation, fig.width = 8.75, fig.height = 6.25}
ggplot(enrollment, aes(type, students, fill = type)) +
  geom_col() +
  scale_fill_aib() +
  scale_y_continuous(labels = aib_label("comma")) +
  labs(
    title = "U.S. K\u201312 Enrollment by Sector",
    x = NULL,
    y = "Students"
  ) +
  theme_aib(base_size = 16) +
  theme(legend.position = "none")
```

## Recommended settings for widescreen slides

Most presentation software uses a 16:9 aspect ratio. A good starting
point is to export at 10 by 5.625 inches (or the proportional 8.75 by
6.25 for a 1.4:1 ratio that leaves room for slide titles):
```r
ggsave(
  "figure.png",
  width = 10,
  height = 5.625,
  dpi = 300
)
```

## Using gridlines for data-dense slides

If your slide includes a time-series or scatter plot where readers need
to trace values, use `theme_aib_grid()` with an increased `base_size`:

```{r grid-slide, fig.width = 8.75, fig.height = 6.25}
set.seed(42)
years <- 2005:2023
gap_data <- data.frame(
  year = rep(years, 2),
  group = rep(c("Higher-Income", "Lower-Income"), each = length(years)),
  score = c(
    270 + cumsum(rnorm(length(years), 0.3, 0.8)),
    240 + cumsum(rnorm(length(years), 0.8, 0.9))
  )
)

ggplot(gap_data, aes(year, score, color = group)) +
  geom_line(linewidth = 1) +
  scale_color_aib() +
  scale_x_continuous(breaks = seq(2005, 2025, 3)) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Reading Achievement Gap Over Time",
    subtitle = "4th-grade scores by household income",
    x = NULL,
    y = "Average Reading Score"
  ) +
  theme_aib_grid(base_size = 16)
```

## Quick reference

| Context | Theme call |
|---|---|
| Print / report | `theme_aib()` |
| Slides (clean) | `theme_aib(base_size = 16)` |
| Slides (gridlines) | `theme_aib_grid(base_size = 16)` |
| Slides (map) | `theme_aib_map(base_size = 16)` |
