---
title: "Themes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  fig.width = 7,
  fig.height = 5,
  dpi = 600
)
```

```{r library}
library(ggaib)
library(ggplot2)
```

The ggaib package includes three theme variants. All share the same fonts
and color conventions but differ in gridline and axis treatment to suit
different contexts.

## `theme_aib()` --- Publication

The default theme targets journal articles and reports. It uses a clean,
minimal design with no gridlines, thin axis lines, and a bottom legend.

```{r theme-pub}
set.seed(42)
districts <- data.frame(
  spending = c(rnorm(40, 11, 2), rnorm(40, 15, 2.5)),
  avg_score = c(rnorm(40, 250, 15), rnorm(40, 270, 12)),
  type = rep(c("Urban", "Suburban"), each = 40)
)

ggplot(districts, aes(spending, avg_score, color = type)) +
  geom_point(size = 2) +
  scale_color_aib() +
  scale_x_continuous(labels = aib_label("dollar")) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Per-Pupil Spending and Math Scores",
    subtitle = "Simulated school district data",
    x = "Per-Pupil Expenditures ($1,000s)",
    y = "Average Math Score",
    caption = "Note: Simulated data for illustration"
  ) +
  theme_aib()
```

### Gridlines

All theme variants accept a `gridlines` argument. Use `"x"`, `"y"`, or
`"xy"` to add light gray major gridlines:

```{r gridlines}
ggplot(districts, aes(spending, avg_score, color = type)) +
  geom_point(size = 2) +
  scale_color_aib() +
  scale_x_continuous(labels = aib_label("dollar")) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Adding Horizontal Gridlines",
    x = "Per-Pupil Expenditures ($1,000s)",
    y = "Average Math Score"
  ) +
  theme_aib(gridlines = "y")
```

## `theme_aib_grid()` --- Data-Dense

Identical to `theme_aib()` but defaults to `gridlines = "xy"`. Use
this for time-series or scatter plots where gridlines help readers trace
values.

```{r theme-grid}
set.seed(42)
years <- 2005:2023
gap_data <- data.frame(
  year = rep(years, 2),
  group = rep(c("Higher-Income", "Lower-Income"), each = length(years)),
  score = c(
    270 + cumsum(rnorm(length(years), 0.3, 0.8)),
    240 + cumsum(rnorm(length(years), 0.8, 0.9))
  )
)

ggplot(gap_data, aes(year, score, color = group)) +
  geom_line(linewidth = 1) +
  scale_color_aib() +
  scale_x_continuous(breaks = seq(2005, 2025, 3)) +
  scale_y_continuous(limits = c(215, 300), breaks = seq(200, 300, 20)) +
  labs(
    title = "Reading Achievement Gap Over Time",
    subtitle = "4th-grade scores by household income",
    x = NULL,
    y = "Average Reading Score"
  ) +
  theme_aib_grid()
```

## `theme_aib_map()` --- Maps

Removes axis lines, ticks, labels, and titles. Retains the plot title,
subtitle, caption, and legend styling.

```{r theme-map}
set.seed(42)
states <- ggplot2::map_data("state")
spending_by_state <- data.frame(
  region = unique(states$region),
  spending = runif(length(unique(states$region)), 7, 24)
)
map_df <- merge(states, spending_by_state, by = "region")

ggplot(map_df, aes(long, lat, group = group, fill = spending)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  scale_fill_aib_c() +
  labs(
    title = "Per-Pupil Spending by State",
    fill = "$ (1,000s)",
    caption = "Note: Simulated data for illustration"
  ) +
  coord_fixed(1.3) +
  theme_aib_map()
```

## Common parameters

All three themes accept the same arguments:
- `base_size` --- Base font size in points (default 11). Increase to
  16 or higher for presentations (see `vignette("presentations")`).
- `base_family` --- Override the base font family. When `NULL` (the
  default), the registered AIB body font is used.
- `gridlines` --- Which major gridlines to draw: `"none"`, `"x"`,
  `"y"`, or `"xy"`.
